---
title: "Data Analysis-1"
format: html
editor: visual
---

### Introduction

The data described below deals with three different types of flour that were inoculated with E. coli. The goal was to study the survival persistence of E. coli. Each type of flour was given three different levels of water and the colony-forming-unit (CFU) per gram was measured at ten unequally spaced time points.

### About the data

The data set provided by the client includes 270 observations. The three types of flour are wheat, buck, and cake, with each accounting for 90 of the observations. These are recorded in the categorical variable *Flour*. The variable *aw* measures the water activity with the three water levels 0.45, 0.55 and 0.65 units. *Day* and *REP* record the day and repetition respectively. The data set also includes the log of CFU/g instead of the untransformed values. This variable is titled *Ecoli_Load* in the data set.

#### Distribution of E.coli Measurements

The first information to consider is the apparent distribution of our dependent variable, which in this case will be the log of CFU/g. To use a linear model or linear mixed model, the independent variable should be approximately normally distributed. We can check this assumption with two plots.

The first plot is a histogram. If the data follows an approximate normal distribution, the histogram should follow a bell shape. The histogram below, shows signs of a mild left skew, but this does not mean the data is not approximately normal.

```{r}
#| echo: false
Data_1 <- read.csv("Data-1.csv")
hist(Data_1$Ecoli_Load)
qqnorm(Data_1$Ecoli_Load)
qqline(Data_1$Ecoli_Load, col = "red")


```

The second plot to consider is the qq plot above. A QQ plot uses theoretical values from a normal distribution as the x-axis, and the observations from the data on the y-axis. If the data is normally distributed, most of the points should fall on a 45 degree line in the positive direction. In our plot we can see that the majority of the points do fall close to this line. There is some values further from the line at both tails, but this is not uncommon.

Looking at both the histogram and the qq plot, the assumption that the log of the E.coli measurements are approximately normally distributed is not clearly violated. Therefore, a linear model or mixed model will be used later, as opposed to the generalized version of either.

#### Impact of Flour Type

One key question in this study is what effect the flour type has on the E.coli levels. There are some differences in the log of CFU/g we can get from the data set right away as it relates to flour. Buck flour has a mean log CFU/g of roughly 5.49, a median of 5.59, and a standard deviation of 0.85. Cake flour has a mean of 5.53, a median of 5.91, and a standard deviation of 1.73. The same three metrics for wheat flour are roughly 4.61, 4.67, and 1.27 respectively. These differences will become more apparent in some of the graphs below.

```{r}
#| echo: false
# Load the dataset first

Data_1 <- read.csv("Data-1.csv")


suppressPackageStartupMessages(library(dplyr))
#Data_1 %>%
#  group_by(Flour) %>%
#  summarise(
#    Mean_Ecoli_Load = mean(Ecoli_Load, na.rm = TRUE),
#    Median_Ecoli_Load = median(Ecoli_Load, na.rm = TRUE),
#    SD_Ecoli_Load = sd(Ecoli_Load, na.rm = TRUE)
#  )
boxplot(Ecoli_Load ~ Flour, data = Data_1, main="Log(E. coli Load) by Flour", xlab="Flour", ylab="Log(E. coli Load)")

```

To get an idea of this graphically, let's consider the box plot above. We can see the interquartile range (IQR) of the three types of wheat vary, with cake flour being the most spread out followed by wheat flour. These differences match the standard deviations discussed above. What stands out about in the box plot are a handful of outliers at the lower end of the box plot for the cake flour observations. This suggests that some samples had much lower E. coli loads compared to the bulk of the data. Neither of the other two types show any outliers.

From both the plot above and the metrics discussed earlier we can say cake flour tends to have the highest E. coli load, followed by buck flour, then wheat flour. This indicates that the type of flour may play a role in the persistence of E. coli.

To see whether the suspected relationship between the flour type and the persistence of E. coli is significant, we can view an ANOVA table. The ANOVA table below is based on a simple linear model, which follows the form \$\$

log((CFU/g)\_{i})=\beta\_0 + \beta\_1 \text{Flour}*i +* \epsilon{i}

\$\$ where $\beta_0$ represents the y-intercept, $\beta_1$ represents the coefficient to the type of flour, for the three flower types which are denoted by $i$. \epsilon\_{i} represents the error. The results of this model can be seen below.

```{r}
#| echo: false
aov_results <- aov(Ecoli_Load ~ Flour, data = Data_1,)
summary(aov_results)

```

We can see the relationship between the type of flour and the log of CFU/g of E.coli is highly significant. Now we will look at contrasts based on the type of flour. Since we are comparing the effect three different flour types have on the E.coli measurement, we should use Tukey's Honest Significant Difference (HSD) test. This method is less likely to result in a false positive, also called a type I error, when comparing multiple treatments.

(https://www.blackwellpublishing.com/specialarticles/jcn_8_304.pdf)

```{r}
#| echo: false
TukeyHSD(aov_results)

```

Tukey's HSD test indicates significant differences in our E. coli measurement between wheat and both buck and cake flour types. There is no significant difference between cake and buck flour types.

#### Other variables

While we have created a simple model, including more information may be an improvement. As discussed previously, the data comes with three other variable. Among the other variables, knowing effect the water activity and time have on the E.coli levels is important.

#### Water Activity

First let's consider the effect water activity has on both the E.coli levels and the types of flour. This can be done through an interaction plot.

```{r}
#| echo: false
interaction.plot(x.factor = Data_1$aw,
                 trace.factor = Data_1$Flour,
                 response = Data_1$Ecoli_Load,
                 type = "b",
                 legend = TRUE,
                 xlab = "Water Activity",
                 ylab = "Log(E. coli Load)",
                 trace.label = "Flour Type",
                 col = c("blue", "red", "green"),
                 pch = 19)

```

The interaction plot above shows that higher water levels seem to increase the E.coli measurement among the buck flour observations while decreasing the measurement among the cake flour observations. The wheat flour observations have little change according to the graph. To answer the question of whether this is significant or not, a new model could be fit with water activity added, but first let's consider another variable.

#### Day of Measurement

Knowing how E.coli levels change over time also seems like important information to consider. Below is a plot of our E.coli measurements over time, which comes from the variable *Day*. There is a trend with the E.coli levels going down as the day of the measurement increases.

```{r}
#| echo: false
plot(x=Data_1$Day, y=Data_1$Ecoli_Load, xlab = "Day", ylab = "Log(Ecoli_Load)", main = "E.Coli Measurments over Time")

```

### ANOVA Tables

Using the variables that were just graphically investigated, let's create a linear model to see the statistical effect *Flour*, *Day*, and *aw* have on the E.coli measurement, which is still in log form.

In the ANOVA below *Flour*, *Day*, and *aw* are considered independent variables, while the E.coli measurement is the dependent variable, as before.

```{r}
#| echo: false
aov_results2 <- aov(Ecoli_Load ~ Flour + Day + aw, data = Data_1,)
summary(aov_results2)

```

Above we can see the type of flour is still highly significant, as is the variable *Day*, which makes sense from what we saw in the time plot. The water activity is not significant though, as the p-value is above 0.5. Despite not being significant, the water activity may be important within the context of the study, so it will remain in the model for now.

One thing missing in the ANOVA above is any interaction terms. Below is an ANOVA designed to account for possible interactions between the variables. Both *Flour* and *Day* are still highly significant, and we can see their interaction is as well. While the water activity on its own is not significant, the interaction between it and the type of flour is at a marginal level. Therefore we should keep both *aw* and its interaction with *Flour* in the model, but may consider removing others.

```{r}
#| echo: false
interaction_aov <- aov(Ecoli_Load ~ Flour * Day * aw, data = Data_1)
summary(interaction_aov)

```

The ANOVA results of removing the interaction between water activity and the day along with the interaction of all three are shown below.

```{r}
#| echo: false
interaction_aov2 <- aov(Ecoli_Load ~ Flour * Day + aw * Flour, data = Data_1)
summary(interaction_aov2)

```

### Accounting for Replications

With three models to consider now, the last variable in the data set, *REP*, should be considered. *REP* measures the replications. This variable should be included as a random effect. Doing so allows us account for additional variability. It also means we are now working with a linear mixed model instead of the mixed model above. The resulting ANOVA table from adding the random effect to our third model is below.

```{r}
#| echo: false
suppressPackageStartupMessages(library(lme4))
#install.packages("lmerTest")
suppressPackageStartupMessages(library(lmerTest))

# Mixed-effects model
interaction_mixed2 <- lmer(Ecoli_Load ~ Flour * Day + aw * Flour + (1 | REP), data = Data_1)
anova(interaction_mixed2)

```

In this model, which excludes two of the possible interaction terms but has a random , there is a similar level of significance as above.

In the ANOVA below, which includes terms for all possible interactions as well as, there are fewer values that appear significant. The day is considered highly significant when looking at the corresponding p-value, while the type of flour is now only moderately signifcant.

```{r}
#| echo: false
suppressPackageStartupMessages(library(lme4))
#install.packages("lmerTest")
suppressPackageStartupMessages(library(lmerTest))

# Mixed-effects model
interaction_mixed <- lmer(Ecoli_Load ~ Flour * Day * aw  + (1 | REP), data = Data_1)
anova(interaction_mixed)

```

When comparing the two mixed models that were used in the ANOVA tables, it would seem excluding two of the interaction terms increases the significance of the other variables in the model. Looking at an ANOVA is not all that must be done to queck the model though.

### Checking Assumptions of the Potential Model

Before declaring a model, we should check whether a potential model corresponding to an ANOVA table above violates any assumptions around mixed models.

Let's consider the mixed model that excluded two of the interaction terms first. The four plots below check some of the assumptions that must hold for this model to be used.

The QQ plot and histogram are similar to what was used above, excpet this time residuals are being plotted instead of E.coli levels. One of our assumptions is that the residuals are normally distributed, and by looking at the two plots, this assumption seems to hold.

Another assumption we need to check is whether the residuals have constant variance, also known as homoscedasticity. This can be done through the residuals vs. fitted values plot below. The plot shows little to no trend with the residuals, which is ideal. While there are more data points on the right side of the graph they still appear random, leading to the conclusion that the constant variance assumption holds.

The plot in the bottom right corner measures how the values generated by the model compare to the actual values observed in the data set. Most points falling on a 45 degree line is a good sign, as it seems the model's predictions hold up based on our sample.

```{r}
#| echo: false
# Define observed values from the data
observed_values <- Data_1$Ecoli_Load  # Assuming Data_1 is your data frame and Ecoli_Load is the response variable

fitted_values2 <- fitted(interaction_mixed2)
residuals2 <- residuals(interaction_mixed2)

# Set up the plotting area to have 2 rows and 2 columns
par(mfrow=c(2, 2))
par(mar = c(4, 4, 2, 1))
# Plot 1: Residuals vs. Fitted Values
plot(fitted_values2, residuals2, main="Residuals vs Fitted Values", xlab="Fitted Values", ylab="Residuals")
abline(h=0, col="red")

# Plot 2: Q-Q Plot of Residuals
qqnorm(residuals2, main="Q-Q Plot of Residuals")
qqline(residuals2, col = "red")

# Plot 3: Histogram of Residuals
hist(residuals2, main="Histogram of Residuals", xlab="Residuals", breaks=20)

# Plot 4: Predicted vs. Observed Values
plot(observed_values, fitted_values2, main="Predicted vs Observed Values", xlab="Observed Values", ylab="Predicted Values")
abline(a=0, b=1, col="red")

# Reset plotting layout to default (optional)
par(mfrow=c(1, 1))
```

Since all of the assumptions discussed above hold, the mixed model including only some interactions can be used.

### Linear Mixed Model

After checking the assumptions we can confirm our model is

$$
\begin{align*}
\log\left(\frac{CFU}{g}\right)_{ijk} &= \beta_0 + \beta_1 \cdot \text{Flour}_i + \beta_2 \cdot \text{Day}_j \\
&\quad + \beta_3 \cdot \text{aw}_k + \beta_4 \cdot (\text{Flour}_i \times \text{Day}_j) \\
&\quad + \beta_5 \cdot (\text{aw}_k \times \text{Flour}_i) + u_j + \epsilon_{ijk}
\end{align*}
$$

where $\beta_b$ for $b=1,2,3,4,5$ represent the fixed effect coefficients, $u_{j}$ represents the random effect associated with *REP*, and $\epsilon_{ijk}$ is the error term.

Now it is time to look at the full summary of our model.

```{r}
#| echo: false
coefficients<-
model_summary <- summary(interaction_mixed2)
print(model_summary)

```

Using the estimates above we can rewrite the model as 
$$
\begin{align*}
\log\left(\frac{CFU}{g}\right)_{ijk} &= 5.718458 + 2.717035 \cdot \text{Flour}_{cake} - 0.070328 \cdot \text{Flour}_{wheat} \\
&\quad - 0.019919 \cdot \text{Day} + 0.913333 \cdot \text{aw} \\
&\quad - 0.026779 \cdot (\text{Flour}_{cake} \times \text{Day}) - 0.010729 \cdot (\text{Flour}_{wheat} \times \text{Day}) \\
&\quad - 3.068333 \cdot (\text{Flour}_{cake} \times \text{aw}) - 0.755000 \cdot (\text{Flour}_{wheat} \times \text{aw}) \\
&\quad + u_j + \epsilon_{ijk}
\end{align*}
$$

Note buck flour is not shown.  The reason for this is since the type of flour is a category variable, whenever the type of is buck, all terms involving cake or wheat flour will equal zero, and the remaining terms indicated buck flour.


# Need more comparison stuff here -- WHERE RYAN STOPPED

### Model Comparison

LM model AIC: 500.80 (lower is better) GLMM model AIC: 576.4 The LM model has a lower AIC, indicating a better fit based on this criterion.

BIC:

LM model BIC: 547.58 (lower is better) GLMM model BIC: 626.8 The LM model also has a lower BIC, further suggesting it is the better model by this criterion.

***Conclusion:*** AIC and BIC both indicate that the LM model fits the data better than the GLMM. However, model choice should also consider the assumptions about the data distribution:

```{r}
#| echo: false
# Load the necessary library
suppressPackageStartupMessages(library(lme4))

# Fit the LM model
interaction_mixed <- lmer(Ecoli_Load ~ Flour + Day + aw + aw * Day + Flour * Day * aw + (1 | REP), data = Data_1)

# Extract log-likelihood
logLik_LM <- logLik(interaction_mixed)

# Number of parameters
num_params_LM <- length(fixef(interaction_mixed)) + length(VarCorr(interaction_mixed)$REP)

# Number of observations
n_LM <- nrow(Data_1)

# Calculate AIC and BIC
AIC_LM <- -2 * logLik_LM + 2 * num_params_LM
BIC_LM <- -2 * logLik_LM + num_params_LM * log(n_LM)

# Print AIC and BIC for the LM model
c(AIC_LM , BIC_LM)


```
